### 2013-02-16 doc/icml2013/algo.tex
edit

In this section, we describe our Spectral Experts algorithm
for estimating model parameters $\theta = (\pi, B)$.
The algorithm consists of two steps:
(i) low-rank regression to estimate certain symmetric tensors;
and (ii) tensor factorization to recover the parameters.
The two steps can be performed efficiently using
convex optimization and tensor power method, respectively.
% Motivate how we can recover $B$ from regression by lower order moments
%ideally from lower order moments.
%Conventionally, when using the method of moments, the
%moments of observed variables are expressed algebraically in terms of
%the model parameters. Using these equations, we attempt to solve for the
%model parameters using empirical estimates of the moments.

@@

\begin{algorithm}[t]
  \caption{Spectral Experts}
  \label{algo:spectral-experts}
  \begin{algorithmic}[1]
    \INPUT Three independent datasets $\mathcal{D}_r = \{ (\xn{1}, \yn{1}), \cdots, (\xn{n}, \yn{n}) \}$ for $r = 1, 2, 3$;
    regularization strengths $\lambda_n^{(2)} = \frac{c_2}{\sqrt{n}}$, $\lambda_n^{(3)} = \frac{c_3}{\sqrt{n}}$;
    observation noise variance $\sigma^2$.
    \OUTPUT Parameters $\hat\theta = (\hat \pi, [\hat \beta_1 \mid \cdots \mid \hat \beta_k])$.
    \STATE Estimate compound parameters $M_2, M_3$ using \textbf{low-rank regression}:
    \begin{align}
      &\hat M_1 = \arg\min_{M_1} \label{eqn:estimateM1} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_1} (\innerp{M_1}{x} - y)^2, \nonumber \\
      &\hat M_2 = \arg\min_{M_2} \quad \lambda_n^{(2)} \|M_2\|_* + \label{eqn:estimateM2} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_2} (\innerp{M_2}{x\tp{2}} + \sigma^2 - y^2)^2, \nonumber \\
      &\hat M_3 = \arg\min_{M_3} \quad \lambda_n^{(3)} \|M_3\|_* + \label{eqn:estimateM3} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_3} (\innerp{M_3}{x\tp{3}} + 3 \sigma^2\innerp{\hat M_1}{x} - y^3)^2. \nonumber
    \end{align}
    \STATE Estimate the parameters $\theta = (\pi, B)$ using \textbf{tensor factorization}:
    \begin{enumerate}
      \item [(a)] Compute whitening matrix $\hat W \in \Re^{d \times k}$ (such that $\hat W^\top
      \hat M_2 \hat W = I$) using SVD.
      \item [(b)] Compute the eigenvalues $\{\hat a_h\}_{h=1}^k$
      and eigenvectors $\{\hat v_h\}_{h=1}^k$
      of the whitened tensor $\hat M_3(\hat W, \hat W, \hat W) \in \Re^{k \times k \times k}$
      by using the robust tensor power method.
    \item [(c)] Set parameter estimates $\hat\pi_h = \frac{1}{\hat a_h^2}$
    and $\hat\beta_h = (\hat W^{\top})^\dagger (\hat a_h \hat v_h)$.
    \end{enumerate}
  \end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Convergence of Spectral Experts]
  \label{thm:convergence}
Each dataset $\sD_p$ (for $p = 1, 2, 3$) consists of $n$ i.i.d. points drawn from the mixture
of linear regressions model with parameter $\theta^*$
(having three independent copies simplifies the analysis).
Let $\|x\|_2 \le R$ with probability $1$
and let $\|\beta_h^*\|_2 \le L$ for all $h \in [k]$.
Let $\Sigma_p \eqdef \E[\cvec(x\tp{p})\tp{2}]$, 
and assume $\Sigma_p \succ 0$ for each $p \in \{1,2,3\}$.
Assume $d \ge k$.
Suppose the number of samples is
$n = \max(n_1,n_2)$
where $n_1 = \Omega(d^p R^p \sigmamin(\Sigma_p)^{-1} \sqrt{p \log(d) \log(1/\delta)})$ and
$n_2 = \Omega\left( \sigma^6 L^6 R^{12} k \sigmamin(\Sigma_p)^{-2} \frac{\log^3 (1/\delta)}{\epsilon^2} \right)$.
If the regularization strengths $\lambda_n^{(2)}$ and $\lambda_n^{(3)}$ are
set to $\Omega(\sigma^3 L^3 R^6 \sqrt{\frac{\log(1/\delta)}{n}})$,
then the parameter estimates $\hat\theta = (\hat\pi, \hat B)$ returned by
\algorithmref{algo:spectral-experts} (with the columns appropriate permuted)
satisfies 
$\|\hat\pi - \pi^*\|_{\infty} \le \epsilon$
and for all $h \in [k]$,
$\|\hat\beta_h - \beta^*_h\|_2 \le \frac{\epsilon}{\pi_h^*}$.
\end{theorem}

\begin{proof}
By \reflem{lowRank} along with \reflem{lowRankLower} and \reflem{lowRankUpper} (discussed below),
we can control the Frobenius norm of the parameter estimates:
If $n \ge n_1$, then
\begin{align}
  \|\hat M_p - M_p\|_F = O\left( \sigma^3 L^3 R^6 k^{\frac12} \sigmamin(\Sigma_p)^{-1} \sqrt{\frac{\log^3 (1/\delta)}{n}} \right).
  %\|\hat M_p - M_p\|_F = O\left( \frac{\sigma^3 L^3 R^6 \sqrt{k} \log^3 (1/\delta)}{\sqrt{n} (\sigmamin(\Sigma_p) - d^p R^p \sqrt{\frac{p \log(d) \log(1/\delta)}{n}}} \right).
\end{align}
The Frobenius norm upper bounds the operator norm,
so we can directly apply Theorem 5.1 of \cite{AnandkumarGeHsu2012}.
%Let $T^* = M_3^*(W^*, W^*, W^*)$.
%Need to convert error from $\epsilon$
%hit by $(W^\top)^\dagger$
\end{proof}

@@

We note that the positive definite constraints are more stringent.
For example, suppose $x = (1, t, t^2)$,
the common polynomial basis expansion, so that all the coordinates are deterministically related.
Then $\E[x\tp{2}] \succ 0$ can easily be satisfied, say with $x \sim \normal{0}{I}$.
However, one can verify that $\E[\cvec(x\tp{2})\tp{2}]$ is singular,
since $\cvec(x\tp{2}) = [1 \cdot 1, t\cdot t, 2(1 \cdot t^2), 2(t \cdot t^2), (t^2 \cdot t^2)]$ contains
components $t \cdot t$ and $2(1 \cdot t^2)$, which are linearly dependent.
So our method would not be able to identify the parameters of a mixture of
linear regressions for this distribution over $x$.

We now show that some amount of unidentifiability is intrinsic to the model,
not just our estimation procedure.
Suppose $x = (t, \dots, t^d)$.
Even if we observed all moments
$\E[x\tp{p}y\tp{p}]$ and $\E[x\tp{2p}]$ for $p \in [r]$,
all the coordinates would be monomials up to degree $2dr$,
the therefore the moments live in a $2(dr)$-dimensional subspace.
The parameters live in a subspace of at least dimension $dk$.
Therefore, at least $r \ge k/2$ moments are required for identifiability of any
algorithm for this monomial example.

%In fact, considering moments up to power $p
%Regressing on all powers up to $p$ yield at most $3$.

%$1, x_1, x_1^2, \dots, x_1^p$ is
%neither quadratically nor cubically independent because the product
%space is strictly a subset of $\Re^{\frac{d (d-1)}{2}}$ since it
%contains $x_1 \times x_1 = x_1^2 \times 1$, and $x_1 \times x_1 \times
%x_1 = x_1^2 \times x_1 \times 1 = x_1^3 \times 1 \times 1$.
%In fact, the
%number of unique terms in $x\tp{p}$ is only of the order of $p \times
%d$, and hence we will not have enough equations to solve for $M_2$ or
%$M_3$ unless we consider $p = d^2$ moments.

%The mixture of linear regressions model is identifiable subject to
%permutations of the components. However, our approach
%employs regression on $x\tp{2}$ and $x\tp{3}$, and thus requires
%$\mathcal{X} \otimes \mathcal{X}$ and $\mathcal{X} \otimes \mathcal{X}
%\otimes \mathcal{X}$ to be linearly independent modulo symmetry. We call
%these condition {\em quadratic} and {\em cubic} independence.

%This condition trivially holds when the space $\mathcal{X}$ consists of
%independent vectors $\{ x_1, x_2, \cdots, x_d \}$. Unfortunately, it
%places some constraints on non-linear featurizations.

% \subsection{Recovering $M_2$ and $M_3$}

% In the case of the mixture of linear regressions, we have two observed
% quantities, $x$ and $y$. Unfortunately, because $y$ is a scalar
% quantity, we have only one equation for every moment of $y$, and with
% this naive approach, would have to consider $O( (k+1) D)$ moments to
% recover $B \in \Re{k \times D}$ and $\pi \in \Re^k$.

% To keep the discussion simple, we will consider the case where there is
% no noise, i.e.  $\sigma^2 = 0$. In this case, the moments of $y$ are,
% \begin{eqnarray*}
%   \E[ y ] &=& (\sum_{k=1}^{k} \pi_{k} \beta_k)^T x \\
%           &=& M_1^T \E[x] \\
%   \E[ y^2 ] &=& \sum_{k=1}^{k} \pi_{k} \E[(\beta_k^T x)]^2 \\
%   &=& \sum_{k=1}^{k} \pi_{k} (\beta_k\tp{2} \odot \E[x\tp{2}]) \\
%   &=& M_2 \odot \E[x\tp{2}] \\
%   \E[ y^p ] &=& M_p \odot \E[x\tp{p}].
% \end{eqnarray*}
% 
% We note that joint moments with $x$ will not help either; consider
% \begin{eqnarray}
%   \E[ y^p x^q ] &=& M_p \odot E[ x\tp{p+q} ].
% \end{eqnarray}
% The elements of $\E[ y^p x^q ]$ are linear combinations of $\E[y^p]$,
% and hence we do not get any new equations; \todo{This isn't exactly
% correct; should we even be presenting this line of reasoning?}. This is
% to be expected, since any information about the $\beta$s is really
% encoded in $y$, and not in the $X$.

>>>>

In this section, we describe our Spectral Experts algorithm
for estimating model parameters $\theta = (\pi, B)$.
The algorithm consists of two steps:
(i) low-rank regression to estimate certain symmetric tensors;
and (ii) tensor factorization to recover the parameters.
The two steps can be performed efficiently using
convex optimization and tensor power method, respectively.

@@

\begin{algorithm}[t]
  \caption{Spectral Experts}
  \label{algo:spectral-experts}
  \begin{algorithmic}[1]
    \INPUT Three independent datasets $\mathcal{D}_r = \{ (\xn{1}, \yn{1}), \cdots, (\xn{n}, \yn{n}) \}$ for $r = 1, 2, 3$;
    %regularization strengths $\lambda_n^{(2)} = \frac{c_2}{\sqrt{n}}$, $\lambda_n^{(3)} = \frac{c_3}{\sqrt{n}}$;
    regularization strengths $\lambda_n^{(2)}$, $\lambda_n^{(3)}$;
    observation noise variance $\sigma^2$.
    \OUTPUT Parameters $\hat\theta = (\hat \pi, [\hat \beta_1 \mid \cdots \mid \hat \beta_k])$.
    \STATE Estimate compound parameters $M_2, M_3$ using \textbf{low-rank regression}:
    \begin{align}
      &\hat M_1 = \arg\min_{M_1} \label{eqn:estimateM1} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_1} (\innerp{M_1}{x} - y)^2, \nonumber \\
      &\hat M_2 = \arg\min_{M_2} \quad \lambda_n^{(2)} \|M_2\|_* + \label{eqn:estimateM2} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_2} (\innerp{M_2}{x\tp{2}} + \sigma^2 - y^2)^2, \nonumber \\
      &\hat M_3 = \arg\min_{M_3} \quad \lambda_n^{(3)} \|M_3\|_* + \label{eqn:estimateM3} \\
      &\quad\quad\frac{1}{2n} \sum_{(x,y) \in \sD_3} (\innerp{M_3}{x\tp{3}} + 3 \sigma^2\innerp{\hat M_1}{x} - y^3)^2. \nonumber
    \end{align}
    \STATE Estimate the parameters $\theta = (\pi, B)$ using \textbf{tensor factorization}:
    \begin{enumerate}
      \item [(a)] Compute whitening matrix $\hat W \in \Re^{d \times k}$ (such that $\hat W^\top
      \hat M_2 \hat W = I$) using SVD.
      \item [(b)] Compute the eigenvalues $\{\hat a_h\}_{h=1}^k$
      and eigenvectors $\{\hat v_h\}_{h=1}^k$
      of the whitened tensor $\hat M_3(\hat W, \hat W, \hat W) \in \Re^{k \times k \times k}$
      by using the robust tensor power method.
    \item [(c)] Set parameter estimates $\hat\pi_h = \frac{1}{\hat a_h^2}$
    and $\hat\beta_h = (\hat W^{\top})^\dagger (\hat a_h \hat v_h)$.
    \end{enumerate}
  \end{algorithmic}
\end{algorithm}

@@

\begin{theorem}[Convergence of Spectral Experts]
\label{thm:convergence}
Each dataset $\sD_p$ (for $p = 1, 2, 3$) consists of $n$ i.i.d. points drawn from a mixture
of linear regressions model with parameter $\theta^*$
(having three independent copies simplifies the analysis).
Let $\|x\|_2 \le R$ with probability $1$
and let $\|\beta_h^*\|_2 \le L$ for all $h \in [k]$.
Let $\Sigma_p \eqdef \E[\cvec(x\tp{p})\tp{2}]$, 
and assume $\Sigma_p \succ 0$ for each $p \in \{1,2,3\}$.
Assume $d \ge k$.
Suppose the number of samples is
$n = \max(n_1,n_2)$
where $n_1 = \Omega(d^p R^p \sigmamin(\Sigma_p)^{-1} \sqrt{p \log(d) \log(1/\delta)})$ and
$n_2 = \Omega\left( \sigma^6 L^6 R^{12} k \sigmamin(\Sigma_p)^{-2} \frac{\log^3 (1/\delta)}{\epsilon^2} \right)$.
If the regularization strengths $\lambda_n^{(2)}$ and $\lambda_n^{(3)}$ are
set to $\Omega(\sigma^3 L^3 R^6 \sqrt{\frac{\log(1/\delta)}{n}})$,
then the parameter estimates $\hat\theta = (\hat\pi, \hat B)$ returned by
\algorithmref{algo:spectral-experts} (with the columns appropriate permuted)
satisfies 
$\|\hat\pi - \pi^*\|_{\infty} \le \epsilon$
and for all $h \in [k]$,
$\|\hat\beta_h - \beta^*_h\|_2 \le \frac{\epsilon}{\sqrt{\pi_h^*}}$.
\end{theorem}

The proof of the theorem has two parts.
First, we bound the error in the compound parameters estimates $\hat M_2,\hat M_3$
using results from \citet{Tomioka2011}.
Then, we use results from \cite{AnandkumarGeHsu2012} to convert this error
into a bound on the actual parameter estimates $\hat\theta = (\hat\pi, \hat B)$
derived from the robust tensor power method.
But first, let us study a more basic property: identifiability.

@@

We note that the positive definite constraints are more stringent.
For example, suppose $x = (1, t, t^2)$,
the common polynomial basis expansion, so that all the coordinates are deterministically related.
Then $\E[x\tp{2}] \succ 0$ can easily be satisfied, say with $x \sim \normal{0}{I}$.
However, one can verify that $\E[\cvec(x\tp{2})\tp{2}]$ is singular,
since $\cvec(x\tp{2}) = [1 \cdot 1, t\cdot t, 2(1 \cdot t^2), 2(t \cdot t^2), (t^2 \cdot t^2)]$ contains
components $t \cdot t$ and $2(1 \cdot t^2)$, which are linearly dependent.
Therefore, Spectral Experts would not be able to identify the parameters of a mixture of
linear regressions for this distribution over $x$.

We now show that some amount of unidentifiability is intrinsic to estimation from low-order moments,
not just an artifact of our estimation procedure.
Suppose $x = (t, \dots, t^d)$.
Even if we observed all moments
$\E[x\tp{p}y\tp{p}]$ and $\E[x\tp{2p}]$ for $p \in [r]$,
all the resulting coordinates would be monomials up to degree $2dr$,
and thus the moments live in a $2(dr)$-dimensional subspace.
On the other hand, the parameters $\theta$ live in a subspace of at least dimension $dk$.
Therefore, at least $r \ge k/2$ moments are required for identifiability of any
algorithm for this monomial example.

### 2013-02-16 doc/icml2013/regression.tex
edit

\paragraph{Low-rank Regression}
%\label{sec:regression}

@@

>>>>

\paragraph{Low-rank regression}
%\label{sec:regression}

@@

With these lemmas in hand, we now give the proof of \refthm{convergence}.
\begin{proof}
By \reflem{lowRank} along with \reflem{lowRankLower} and \reflem{lowRankUpper} (discussed below),
we can control the Frobenius norm of the parameter estimates:
If $n \ge n_1$, then
\begin{align}
  \|\hat M_p - M_p\|_F = O\left( \sigma^3 L^3 R^6 k^{\frac12} \sigmamin(\Sigma_p)^{-1} \sqrt{\frac{\log^3 (1/\delta)}{n}} \right).
  %\|\hat M_p - M_p\|_F = O\left( \frac{\sigma^3 L^3 R^6 \sqrt{k} \log^3 (1/\delta)}{\sqrt{n} (\sigmamin(\Sigma_p) - d^p R^p \sqrt{\frac{p \log(d) \log(1/\delta)}{n}}} \right).
\end{align}
The Frobenius norm upper bounds the operator norm,
so we can directly apply Theorem 5.1 of \cite{AnandkumarGeHsu2012}.
%Let $T^* = M_3^*(W^*, W^*, W^*)$.
%Need to convert error from $\epsilon$
%hit by $(W^\top)^\dagger$
\end{proof}
